---
title: "Case Study 1_How does a bike-share navigate speedy success"
author: "mohan"
date: "2025-04-08"
output: html_document
---

---
title: "Case Study 1: How Does a Bike-Share Navigate Speedy Success"
author: "Mohan"
date: "2025-04-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Install and Load Packages

```{r packages}
install.packages("tidyverse")
library(tidyverse)

install.packages("conflicted")
library(conflicted)
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")
```

## Load and Inspect Data

```{r load-data}
q1_2019 <- read_csv("Divvy_Trips_2019_Q1.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")

# Rename columns in q1_2019 to match q1_2020
q1_2019 <- rename(q1_2019,
                  ride_id = trip_id,
                  rideable_type = bikeid,
                  started_at = start_time,
                  ended_at = end_time,
                  start_station_name = from_station_name,
                  start_station_id = from_station_id,
                  end_station_name = to_station_name,
                  end_station_id = to_station_id,
                  member_casual = usertype)
```

## Merge and Clean Data

```{r merge-clean}
# Standardize column types
q1_2019 <- mutate(q1_2019,
                  ride_id = as.character(ride_id),
                  rideable_type = as.character(rideable_type))

# Combine datasets
all_trips <- bind_rows(q1_2019, q1_2020)

# Remove unnecessary columns
all_trips <- all_trips %>% 
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, tripduration))

# Recode member types
all_trips <- all_trips %>% 
  mutate(member_casual = recode(member_casual, "Subscriber" = "member", "Customer" = "casual"))
```

## Add Time Columns

```{r datetime}
all_trips <- all_trips %>% 
  mutate(date = as.Date(started_at),
         month = format(date, "%m"),
         day = format(date, "%d"),
         year = format(date, "%Y"),
         day_of_week = format(date, "%A"),
         ride_length = as.numeric(difftime(ended_at, started_at)))
```

## Clean Ride Data

```{r ride-clean}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length < 0), ]
```

## Descriptive Statistics

```{r statistics}
summary(all_trips_v2$ride_length)

aggregate(ride_length ~ member_casual, data = all_trips_v2, FUN = mean)
aggregate(ride_length ~ member_casual + day_of_week, data = all_trips_v2, FUN = mean)
```

## Weekly Ride Patterns

```{r weekday-patterns}
library(lubridate)

all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, 
                                    levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)
```

## Visualization: Number of Rides by Weekday

```{r plot1}
library(ggplot2)
library(scales)

all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()) %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +
  labs(title = "Number of Rides by Rider Type and Weekday", x = "Day of the Week", y = "Number of Rides") +
  theme_minimal()
```

## Visualization: Average Ride Duration by Weekday

```{r plot2}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(average_duration = mean(ride_length)) %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Average Ride Duration by Rider Type and Weekday", x = "Day of the Week", y = "Average Duration (Minutes)") +
  theme_minimal()
```

## Save Processed Data

```{r export-csv}
counts <- aggregate(ride_length ~ member_casual + day_of_week, data = all_trips_v2, FUN = mean)
write.csv(counts, file = "avg_ride_length.csv")
write.csv(all_trips_v2, file = "all_trips_v2.csv")
```

## Pie Chart: Distribution by Rider Type

```{r piechart}
all_trips_v2 %>%
  group_by(member_casual) %>%
  summarise(number_of_rides = n()) %>%
  ggplot(aes(x = "", y = number_of_rides, fill = member_casual)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  labs(title = "Distribution of Rides by Rider Type") +
  theme_minimal()
```

## Line Graph: Avg Ride Duration Over Time

```{r time-series}
library(lubridate)

all_trips_v2 %>%
  mutate(month = floor_date(started_at, "month")) %>%
  group_by(month) %>%
  summarise(average_duration = mean(ride_length, na.rm = TRUE)) %>%
  ggplot(aes(x = month, y = average_duration)) +
  geom_line(color = "blue", size = 1) +
  labs(title = "Average Ride Duration Over Time", x = "Month", y = "Average Duration (Minutes)") +
  theme_minimal()
```

## Boxplot: Ride Duration by Rider Type

```{r boxplot}
all_trips_v2 %>%
  ggplot(aes(x = member_casual, y = ride_length, fill = member_casual)) +
  geom_boxplot() +
  labs(title = "Ride Duration by Rider Type", x = "Rider Type", y = "Ride Duration (Minutes)") +
  theme_minimal()

```
